---
title: "ST793 Final Project"
author: "Emmett Kendall, Neil Dey"
date: "11/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# _Bootstrap Methods: Another look at the Jackknife_
### B. Efron

---

## Introduction 
The format of this blog post will be as follows. In [Section I](#sec1) we begin with a summary and clear description of the main topics discussed in Efron's paper. In [Section II](#sec2), we will include the proofs of major theorems and remarks made. Lastly, in [Section III](#sec3), we will include illustrative examples and simulations that help the reader understand the impact and importance of Bootstrap.

```{r cars}
summary(cars)
```

## Section I: Summary of article {#sec1}

You can also embed plots, for example: 

```{r pressure, echo=FALSE}
plot(pressure)
```

[comment]: <> (Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.)

## Section II: Theory {#sec2}

## Section III: Examples {#sec3}
